spinifex: A Package for Manual Control of Dynamic Linear Projections of Multivariate Data

### 2019-04-18 paper/spinifex-paper.tex
pg1-6 paper edits. grammerly, knit.

\abstract{%
The set of dynamic linear projections of multivariate data collectively
known as \textbf{tours} provides an important tool for extending the
dimensionality of visuals. The R package \textbf{tourr} offers a variety
of path generators and geometric displays for conducting tours. This
paper discusses an extension package, \textbf{spinifex}, that adds
support for the path generation of manual tours and extends the display
of tours to use with the animation packages, \textbf{plotly} and
\textbf{gganimate}. Manual tours are used to explore the sensitivity of
structure as the contributions of a manipulation variable are changed. A
recent paper (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}) visualized the
sensitivity of the hadronic experiments to nucleon structure.
Sensitivity was characterized in non-linear 3D embeddings of the first
10 principal components. This research applies manual tours to this data
showing that manual tours resolve information about the structure of
cluster distribution by bringing to light information that is orthogonal
to static viewing planes.
}

@@

A tour is a multivariate data analysis technique in which a sequence of
linear (orthogonal) projections are viewed as an animation while the
orientation of the projection basis is rotated across time. Each frame
of the sequence corresponds to a small change in the projection for a
smooth transition that preserves continuity. Tours are a crucial tool
extending the dimensionality of visualization which is increasingly
important as datasets and models exist in increasing cardinality. The
package \CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore how sensitive the projection structure is to these
manipulations. As manual tours operate on only one variable at a time,
they particularly useful once a feature of interest has been identified.
One way to identify ``interesting'' features is with the use of a guided
tour (Hurley and Buja
\protect\hyperlink{ref-hurley_analyzing_1990}{1990}). Guided tours
select a very specific path, that which approaches a projection that
optimizes an objective-function, similar to simulated annealing
(Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified an analyst
can then use the ``finer brush'' of the manual tour, by controlling the
contributions of individual variables to explore the sensitivity they
have to the projection structure.

@@

The paper is organized as follows. Section \ref{sec:algorithm} explains
the algorithm using a toy dataset. Section \ref{sec:display} discusses
the display of the animation after the path has been generated. Section
\ref{sec:usage} explains the package functionality and code usage
following the order applied in the algorithm. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to contemporary high energy physics. The last section,
\ref{sec:discussion} summarizes the work and discusses future research.

@@

The steps are described in more detail below with descriptions,
equations, and figures are given inline. Function names are mentioned,
but code usage is discussed in more detail in section \ref{sec:usage}.

@@

\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions.
  \item $\textbf{B} = (B_1,~ B_2)$, any orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate, the manip var.
  \item $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes.
  \item $\phi$, the angle of out-of-projection-plane rotation, coming into the manipulation space.
\end{itemize}

@@

The data is defined. A basis set (ideally orienting to an interesting
feature) must be provided as an initial orientation. One way to identify
a projection containing interesting features is to apply a guided tour.
In a guided tour, the projection sequence is selected by optimizing an
index function via hill-climbing on the projection space. In this case,
the holes index is selected and applied to standardized flea data. The
holes index is maximized when the projected observations are furthest
from the center. Figure \ref{fig:step0} shows a locally optimized
projection of the data. The left panel displays the reference axes of
the projection basis, a visual indication of the magnitude and direction
each variable contributed to the projection. The right panel shows the
data as projected through the basis set described by the reference axes
(left). Data points are colored and given shape according to the species
(while the guided tour was unsupervised with this information).

@@

Call \texttt{view\_basis()} on a basis to produce a \CRANpkg{ggplot2}
(Wickham \protect\hyperlink{ref-wickham_ggplot2:_2016}{2016}) graphic
similar to \textbackslash ref\{fig:step0). Projection space is always
available for display via the matrix multiplication
\(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

@@

In figure \ref{fig:step0} the contributions of the variables tars1 and
aede2 are mostly orthogonal to the contributions of the other four
variables. These two variables explain the variation of the data
distinguishing the purple group from the rest of the sample. Select
aede2 as the manip var, the variable to be manipulated, as it has a
larger contribution to the projection. The question that will be
explored is how important the variable aede2 is to the separation of the
clusters.

@@

Adding this new dimension to our projection space allows for the
coefficients of the manip var to be changed via rotation about the
origin. For example, the ability to lift a piece of paper, rather than
being constrained to a 2D plane. Orthonormalizing rescales the new
vector while the projection down to 2D remains the original basis. Place
the plane horizontally, with the new dimension extending vertically,
with axes projecting back onto the reference axes. Figure
\ref{fig:step2} illustrates this 3D manipulation space with the manip
var highlighted (height of the other variables are not depicted.)

@@

\caption[Manipulation space for controlling the contribution of aede2 of standardized flea data]{Manipulation space for controlling the contribution of aede2 of standardized flea data. Basis selected by a holes-index optimized guided tour lies on the horizontal projection plane, shown in blue. The manip var axis, in red, extends into the vertical manipulation space, allowing the coefficients of the manip var to be changed by rotation around the origin.}\label{fig:step2}
\end{figure}
\end{Schunk}

The representation in figure \ref{fig:step2} can be produced by calling
the function \texttt{view\_manip\_space()}.

@@

For a radial tour fix the (horizontal) angle within the projection
plane, \(\theta\), and define a sequence for the (vertical) angle coming
out of the projection plane, \(\phi\), bringing the initial \(XY\)
contributions of the manip var to a maximum and then to zero before
returning to the initial position. For an oblique manual tour, capture
the user manipulation directly on the projection plane (not the
manipulation space.)

@@

\begin{itemize}
  \item[$c_\theta$] be the cosine of $\theta$
  \item[$c_\phi$]   be the cosine of $\phi$
  \item[$s_\theta$] be the sine of   $\theta$
  \item[$s_\phi$]   be the sine of   $\phi$
\end{itemize}

@@

Figure \ref{fig:step3} illustrates a sequence with 15 projected bases,
showing the reference axes on top the corresponding projected data
points on the below. As a result, we can see that changes in the manip
var controlled the distance between the purple cluster and the remaining
sample, aede2 is crucial in distinguishing this species. Tours are
typically viewed as an animation such a dynamic version of this tour can
be viewed online at
\url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The
page may take a moment to load.

@@

\caption[Radial manual tour manipulating aede2 of standardized flea data]{Radial manual tour manipulating aede2 of standardized flea data.  The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. The change in the projected data shows that aede2 is important for distinguishing the purple species. An animated version can be viewed at  https://nspyrison.netlify.com/thesis/flea\_manualtour\_mvar5/.}\label{fig:step3}
\end{figure}
\end{Schunk}

Animations can be produced using the function
\texttt{play\_manual\_tour()}. This function defaults to an HTML5 widget
produced from \textbf{plotly}. The \texttt{render\_type} argument can be
changed to \texttt{render\_gganimate} for exporting to GIF or MP4 files.

@@

In an appeal to performance, the above operations are performed on the
bases without the use of the larger datasets. After the bases are
brought into the projection-space, however, it is helpful to observe
them with data in the same space. Pre-multiply the data by basis frame
bringing the data into the projection space.

@@

For a 2D scatterplot, plot the first two variables from each frame
statically as in the previous figure, or in sequence, producing an
animated scatterplot. The remaining variable is sometimes linked to a
data point aesthetic (such as size or color) to produce depth cues used
in conjunction with the \(XY\) scatterplot.

@@

In the following section describes and shows usage of the tour related
functions in parallel with the algorithm. The utility functions are
called by the tour animation functions, passed to the arguments
\texttt{render\_type}. Usage is demonstrated below and in the example
documentation.

@@

We'll start by initializing values including a standardized data set
(numeric columns only), a starting basis, a categorical variable for
point aesthetics, and a manip var. To get bearings on the projection,
start by observing the reference axes of the basis with
\texttt{view\_basis()}.

@@

After becoming familiar with this space, select a manip var, the
variable to change the contributions of. Use
\texttt{view\_manip\_space()} View the new space with a dimension
orthogonal to the projection plane where the manip var has a full
contribution. This illustrates how the manip var is manipulated in a
high space not depicted in the manual tour.

@@

Now we are ready to perform a manual tour on the selected variable. Use
\texttt{play\_manual\_tour()} to perform the algorithm as discussed
above, in section \ref{sec:algorithm}. The argument
\texttt{render\_type} can be changed to \texttt{render\_gganimate} for a
respective animation.

@@

Aside from the code used in the algorithm section a previously generated
tour path can also be animated via \texttt{play\_tour\_path()}. Utility
functions can also be passed into arguments: for example animation
format can be changed by setting
\texttt{render\_type\ =\ render\_gganimate}. View the guided tour path
as a gganimate function:

@@

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

>>>>

\abstract{%
The set of dynamic linear projections of multivariate data collectively
known as \textbf{tours} provides an important tool for extending the
dimensionality of visuals. The R package \textbf{tourr} offers a variety
of path generators and geometric displays for conducting tours. This
paper discusses an extension package, \textbf{spinifex}, that adds
support for the path generation of manual tours and extends the display
of tours to use with the animation packages, \textbf{plotly} and
\textbf{gganimate}. Manual tours are used to explore the sensitivity of
structure as the contributions of a manipulation variable are changed. A
recent paper (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}) visualized the
sensitivity of the hadronic experiments to nucleon structure.
Sensitivity was characterized in non-linear 3D embeddings of the first
10 principal components. This research applies manual tours to this data
showing that manual tours resolve information about the structure of
cluster distribution by bringing revealing information that is
orthogonal to static viewing planes.
}

@@

Exploring multivariate spaces is a challenging task and increasingly so
as dimensionality increases. Traditionally, static projections display
an embedding of multivariate data in two dimensions. Such static views
cannot provide depth cues or convey the information of other
orientations. Alternatively, tours are a class of dynamic linear
projections viewed as an animation while the orientation of the
projection is rotated. Tours are a crucial tool extending the
dimensionality of visualization which is increasingly important as
datasets and models exist in increasing cardinality. The package
\CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore their sensitivity to the structure within the projection. As
manual tours operate on only one variable at a time, they particularly
useful once a feature of interest has been identified. One way to
identify ``interesting'' features is with the use of a guided tour
(Hurley and Buja \protect\hyperlink{ref-hurley_analyzing_1990}{1990}).
Guided tours select a very specific path, that which approaches a
projection that optimizes an objective-function, in a manner similar to
simulated annealing (Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified an analyst
can then use the ``finer brush'' of the manual tour, by controlling the
contributions of individual variables to explore the sensitivity they
have on the structure within the projection.

@@

The paper is organized as follows. Section \ref{sec:algorithm} explains
the algorithm using a toy dataset. Section \ref{sec:display} discusses
the display of the animation after the path has been generated. Package
functionality and code usage following the order applied in the
algorithm follows in section \ref{sec:usage}. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to contemporary high energy physics. Then section
\ref{sec:summary} summarizes this paper. Lastly, section
\ref{sec:discussion} discusses possible future research.

@@

These algorithm steps are described in more detail below with
descriptions, equations, and figures are given inline. Function names
are mentioned, but code usage is discussed in more detail in section
\ref{sec:usage}.

@@

\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions
  \item $\textbf{B} = (B_1,~ B_2)$, any orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate, the manip var
  \item $\theta_i$, the sequence of angles of in-projection-plane rotation, for example, on the reference axes
  \item $\phi_i$, the sequence of angles of out-of-projection-plane rotation, coming into the manipulation space
\end{itemize}

@@

The data is defined. A basis set, the orientation of the projection to
explore, must be provided. One way to identify a projection containing
interesting features is to apply a guided tour. In a guided tour, the
projection sequence is selected by optimizing an index function via
hill-climbing on the projection space. In this case, the holes index is
selected and applied to standardized flea data. The holes index is
maximized when the projected observations are furthest from the center.
Figure \ref{fig:step0} shows a locally optimized projection of the data.
The left panel displays the reference axes of the projection basis, a
visual indication of the magnitude and direction each variable
contributes to the projection. The right panel shows the data as
projected through the basis set described by the reference axes (left).
Data points are colored and given shape according to the species (while
the guided tour was unsupervised with this information).

@@

Call \texttt{view\_basis()} on a basis to produce a \CRANpkg{ggplot2}
(Wickham \protect\hyperlink{ref-wickham_ggplot2:_2016}{2016}) graphic
similar to figure \ref{fig:step0}. Projection space is always available
for display via the matrix multiplication
\(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

@@

In figure \ref{fig:step0} the contributions of the variables tars1 and
aede2 are mostly orthogonal to the contributions of the other four
variables. These two variables explain the variation of the data
distinguishing the purple cluster from the rest of the sample. Select
aede2 as the manip var, the variable to be manipulated, as it has a
larger contribution to the projection. The question that will be
explored is how important the variable aede2 is to the separation of the
clusters.

@@

Adding this new dimension to our projection space allows for the
coefficients of the manip var to be changed via rotation about the
origin. For example, the ability to lift a piece of paper, rather than
being constrained to a 2D plane. Orthonormalizing rescales the new
vector, while the projection down to 2D remains as the original basis.
Place the projection plane horizontally, with the new dimension
extending vertically, with axes projecting back onto the reference axes.
Figure \ref{fig:step2} illustrates this 3D manipulation space with the
manip var highlighted (height of the other variables are not depicted).
This representation is produced by calling the
\texttt{view\_manip\_space()} function. Functionally, the manipulation
space is more of a mathematical construct that needn't be visualized but
does help to digest what is happening mathematically.

@@

\caption[Manipulation space for controlling the contribution of aede2 of standardized flea data]{Manipulation space for controlling the contribution of aede2 of standardized flea data. The basis was selected by a holes-index optimized guided tour lies on the horizontal projection plane, shown in blue. The manip var axis, in red, extends into the vertical manipulation space, allowing the coefficients of the manip var to be changed by rotation around the origin.}\label{fig:step2}
\end{figure}
\end{Schunk}

@@

For a radial tour fix the (horizontal) angle within the projection
plane, \(\theta\), and define a sequence for the (vertical) angle coming
out of the projection plane, \(\phi\), bringing the initial \(XY\)
contributions of the manip var to a maximum and then to zero before
returning to the initial position. For an oblique manual tour, capture
the user manipulation directly on the \(XY\) of the projection plane.

@@

\begin{itemize}
  \item[$c_\theta$] be the cosine of $\theta_i$
  \item[$c_\phi$]   be the cosine of $\phi_i$
  \item[$s_\theta$] be the sine of   $\theta_i$
  \item[$s_\phi$]   be the sine of   $\phi_i$
\end{itemize}

@@

Figure \ref{fig:step3} illustrates a sequence with 15 projected bases,
showing the reference axes on the top half with the corresponding
projected data points below. Changes in the basis coefficients of the
manip var correspond to the distance separating the purple cluster and
the remaining sample, aede2 is crucial in distinguishing the variation
of species. Tours are typically viewed as an animation. The animation of
this tour can be viewed online at
\url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The
page may take a moment to load. Animations can be produced using the
function \texttt{play\_manual\_tour()}.

@@

\caption[Radial manual tour manipulating aede2 of standardized flea data]{Radial manual tour manipulating aede2 of standardized flea data.  The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. Manipulating the coefficeients of aede2 changes the distance separating the purple cluster. This shows that aede2 is important for distinguishing this species from the ramaining data. An animated version can be viewed at  https://nspyrison.netlify.com/thesis/flea\_manualtour\_mvar5/.}\label{fig:step3}
\end{figure}
\end{Schunk}

@@

In an appeal to performance, the operations of a tour are performed on
the bases while the larger datasets are not operated on until the
display of data points are needed. After the bases are brought into the
projection-space, however, it is helpful to observe them with projected
data in the same space. Pre-multiply the data by basis frame bringing
the data into the projection space.

@@

For a 2D scatterplot, plot the first two variables within each frame.
Show each frame in sequence to form an animation. The remaining
manipulation space variable can be linked to point aesthetic (such as
size or color) to produce depth cues used in conjunction with the \(XY\)
scatterplot.

@@

In the following section describes and demonstrates usage of the above
functions following the order in the algorithm. The utility functions
are called by the tour animation functions, passed to the arguments
\texttt{render\_type}. Usage is demonstrated below and in the example
documentation.

@@

We'll start by initializing values including a standardized data set
(numeric columns only), a starting basis, a categorical variable for
point aesthetics (optional), and a manip var. To get bearings on the
projection, start by observing the reference axes of the basis with
\texttt{view\_basis()} producing figure \ref{fig:step0}.

@@

After becoming familiar with this space, select a manip var, the
variable to change the contributions of. Use
\texttt{view\_manip\_space()} to view the new space with a dimension
orthogonal to the projection plane where the manip var has a full
contribution. This illustrates how the manip var is manipulated with the
addition of the manipulation space as shown in figure \ref{fig:step2}.

@@

Now we are ready to perform a manual tour on the selected variable. Use
\texttt{play\_manual\_tour()} to perform the algorithm as discussed
above, in section \ref{sec:algorithm}. This is the animated equivalent
of figure \ref{fig:step3}.

@@

This concludes the content of the algorithm section, however, lets cover
animating paths generated in \textbf{tourr}. Animate the previously
generated guided tour path via \texttt{play\_tour\_path()}. Utility
functions can also be passed as arguments into either of the tour
animation function to change the resulting graphics object, set
\texttt{render\_type\ =\ render\_gganimate} to view the animation as a
GIF.

@@

\hypertarget{sec:summary}{%
\subsection{Summary}\label{sec:summary}}
